Nested Stochastic Block Models Applied to the Analysis of Single Cell Data

### 2021-04-19 manuscript.tex
added a section for cell marginals

%\subsection*{Hierarchic model constrains differential gene tests}
%Current approaches to test for gene markers in scRNA-seq data rely on pairwise or one-vs-rest comparisons among cell groups. This strategy assumes that cell groups are independent which, in general, may not be true. A notable exception is the recently introduced treeclimbR \cite{huang_2020} which has been specifically developed to perform differential tests over a tree structure. We reasoned that we could superpose the hierarchical structure inferred by \emph{schist} by fitting a linear model with mixed-effect, so that we model the dependencies using the inferred tree.
\subsection*{Cell affinities can be used for label transfer}

The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme. Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data. We sought to exploit this property to address the task of annotating cells according to a reference sample. To this end we analyzed datasets from \cite{mereu_2020}, which describe mixtures of human PBMC and HEK293T cells profiled with various technologies. We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartzseq 2 and MARS-seq. These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task. After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph. In each merged dataset, we retained cell type annotations for 10XV3 cells, while we assigned a 'Unknown' label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartzseq 2). We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or 'Unknown'. Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood. The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their unannotated label if and only if there is not enough evidence to associate them to another group. We compared the outcome in terms of accuracy to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \emph{ingest}, a tool included in Scanpy also based on \emph{k}NN classification. Analysis of a well defined dataset, such as Quartzseq 2, reveals that the three approaches are equally good in classifying unknown cells (Fig. \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927. When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unknown. This effect is particular evident for \emph{ingest}, in which only CD4 T cells and HEK cells are transfered, resulting in the lowest accuracy (0.243). Conversely, \emph{schist} is able to assign correct labels with higher accuracy (0.641). Moreover, \emph{k}NN methods assign a label to each cell, whereas \emph{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the Unknown state is the most likely). Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity was indeed the appropriate one (Fig. \ref{hist_label_second}).

>>>>

%\subsection*{Hierarchic model constrains differential gene tests}
%Current approaches to test for gene markers in scRNA-seq data rely on pairwise or one-vs-rest comparisons among cell groups. This strategy assumes that cell groups are independent which, in general, may not be true. A notable exception is the recently introduced treeclimbR \cite{huang_2020} which has been specifically developed to perform differential tests over a tree structure. We reasoned that we could superpose the hierarchical structure inferred by \emph{schist} by fitting a linear model with mixed-effect, so that we model the dependencies using the inferred tree.

\subsection*{Cell marginals improve detection of marker genes}

\textcolor{red}{TO BE DONE.}.

\subsection*{Cell affinities can be used for label transfer}

The modelling approach we adopted allows the estimation of the information required to describe a graph given any partitioning scheme. Differences in entropy can be used to perform model selection, that is we can choose which model better describes the data. We sought to exploit this property to address the task of annotating cells according to a reference sample. To this end we analyzed datasets from \cite{mereu_2020}, which describe mixtures of human PBMC and HEK293T cells profiled with various technologies. We chose cells profiled with 10X V3 platform as reference dataset and performed annotation on cells profiled with Quartzseq 2 and MARS-seq. These are at the extremes of the capability to distinguish cell types, so they provide good benchmark configurations for this task. After preprocessing raw data according to the parameters given in \cite{mereu_2020}, we integrated each dataset with 10XV3 into a unified representation using Harmony \cite{Korsunsky_2019}, and computed the \emph{k}NN graph. In each merged dataset, we retained cell type annotations for 10XV3 cells, while we assigned a 'Unknown' label to all cells derived from the other technology (\emph{i.e.} MARS-seq or Quartzseq 2). We then calculated the cell affinity matrix, that is we computed the difference in entropy that can be observed by assigning each cell to each annotation cluster, this being either one of the original cell types or 'Unknown'. Once the matrix has been computed, each cell from the query data is assigned to the group with the highest likelihood. The rationale behind this approach is that if cells belong to the same annotation group, then more information is required to describe the graph if they were annotated as different cell types; hence, cells from the query datasets should retain their unannotated label if and only if there is not enough evidence to associate them to another group. We compared the outcome in terms of accuracy to \emph{k}NN classification, given by the closest entry in the \emph{k}NN graph, and to \texttt{ingest}, a tool included in Scanpy also based on \emph{k}NN classification. Analysis of a well defined dataset, such as Quartzseq 2, reveals that the three approaches are equally good in classifying unknown cells (Fig. \ref{Figure_Label_Transfer}, central column), with accuracies ranging from .870 to .927. When data are noisy, instead, \emph{k}NN-based methods show low accuracy and a tendency to assign the most represented cell group (HEK293T) to the unlabeled cells. This effect is particular evident for \texttt{ingest}, in which only CD4 T cells and HEK cells are transfered, resulting in the lowest accuracy (0.243). Conversely, \texttt{schist} is able to assign correct labels with higher accuracy (0.641). Moreover, \emph{k}NN methods assign a label to each cell, whereas \texttt{schist} does not relabel cells if there are no sufficient evidence (\emph{e.g.} the "Unknown" state is the most likely). Interestingly, we found that for the largest part of cells without assigned label, the second choice by affinity was indeed the appropriate one (Fig. \ref{hist_label_second}).
